---
title: 2. Get started with rbms - collate index
subtitle: Imputing missing counts
author: Reto Schmucki (UKCEH)
date: 28 November 2019
---
---------

From the flight curve and the observed count, we can derive expected values for weeks or days where a site has not been monitored. Together, observed and imputed counts will then be used to compute an abundance index for each site. Site indices can then be use to calculate annual collated indices.

> see 1. Get started with rbms to compute the flight curve object used below (Step 1, 2 and 3)

```{r data}
library(data.table)
library(rbms)
data(m_visit)
data(m_count)
ts_date <- rbms::ts_dwmy_table(InitYear = 2000, LastYear = 2003, WeekDay1 = 'monday')
ts_season <- rbms::ts_monit_season(ts_date, StartMonth = 4, EndMonth = 9, StartDay = 1, EndDay = NULL, CompltSeason = TRUE, Anchor = TRUE, AnchorLength = 2, AnchorLag = 2, TimeUnit = 'w')
ts_season_visit <- rbms::ts_monit_site(m_visit, ts_season)
ts_season_count <- rbms::ts_monit_count_site(ts_season_visit, m_count, sp = 2)
ts_flight_curve <- rbms::flight_curve(ts_season_count, NbrSample = 300, MinVisit = 5, MinOccur = 3, MinNbrSite = 5, MaxTrial = 4, GamFamily = 'nb', SpeedGam = FALSE, CompltSeason = TRUE, SelectYear = NULL, TimeUnit = 'w')
```

##### 4. Impute predicted counts for missing monitoring dates

The `impute_count()` function is using the count data generated from the `ts_season_count()` function and the outcome of the `flight_curve()` function to  and the calculated flight curve (ts_flight_curve$pheno).
To estimate value of missing week, the function will look for the phenology from the nearest year, the extent of the search can be limited by setting the parameter `YearLimit` parameter which by default is not restricted. Like in the previous function, the imputation can be made on a weekly or daily basis (`'w'` or `'d'`).

```{r, impute}
## extract phenology data from the ts_fligh_curve list
pheno <- ts_flight_curve$pheno

impt_counts <- rbms::impute_count(ts_season_count=ts_season_count, ts_flight_curve=pheno, YearLimit= NULL, TimeUnit='w')
```

The `impute_count()` function produces a `data.table` that contains the original `COUNT` values, a series of `IMPUTED_COUNT` over monitoring season, `TOTAL_COUNT` per site and year, `TOTAL_NM` which is the proportion of the flight curve covered by the visits, the `SINDEX` which is the site index and correspond to the sum of both observed and imputed counts over the sampling season.

If the flight curve of a year is missing, the `impute_count()` function will look and use the nearest phenology found. If none is found within the limit of years set by the `YearLimit` parameter, the function will compute no `SINDEX` for that specific year.

```{r, impute_year}
impt_counts_1year <- rbms::impute_count(ts_season_count=ts_season_count, ts_flight_curve=pheno[M_YEAR != 2001, ], YearLimit= 1, TimeUnit='w')

impt_counts_0year <- rbms::impute_count(ts_season_count=ts_season_count, ts_flight_curve=pheno[M_YEAR != 2001, ], YearLimit= 0, TimeUnit='w')
```


##### 5 Site and collated indices

From the imputed count, site index can be extracted with a filter that will only keep the indices that have been monitored of a minimum proportion of the flight curve. Here we set this threshold
to a minimum value of 10% with the `MinFC` parameter.

```{r, sindex}
sindex <- rbms::site_index(butterfly_count = impt_counts, MinFC = 0.10)
```

Using the site indices, annual collated indices can be calculated by fitting a Generalized Linear Model (GLM), where sites and years are set as factors. Here we use the proportion of the flight curve sampled by the observation as a weight for the GLM. To facilitate the fit, we can remove all site where the species was not observed by setting the parameter `rm_zero = TRUE`.
The collated index computed by the `collated_index()` function corresponds to the mean total butterfly count expected on a BMS transect in a given year.

```{r, collated_index}
co_index <- collated_index(data = sindex, s_sp = 2, sindex_value = "SINDEX", glm_weights = TRUE, rm_zero = TRUE)
```

The collated index computed by the `collated_index()` function corresponds to the mean total butterfly count expected on a BMS transect in a given year.

```{r, collated_index_print, echo = FALSE}
co_index
```

This number can then be rescaled on a log(10) scale

```{r log_ind}
co_index <- co_index$col_index
co_index_b <- co_index[COL_INDEX > 0.0001 & COL_INDEX < 100000, ]
co_index_logInd <- co_index_b[BOOTi == 0, .(M_YEAR, COL_INDEX)][, log(COL_INDEX)/log(10), by = M_YEAR][, mean_logInd := mean(V1)]

## merge the mean log index with the full bootstrap dataset
data.table::setnames(co_index_logInd, "V1", "logInd"); setkey(co_index_logInd, M_YEAR); setkey(co_index_b, M_YEAR)
co_index_b <- merge(co_index_b, co_index_logInd, all.x = TRUE)
```

This log scaled indices can then be plotted with the following code, where the average is centred on 2.

```{r, ind_fig, fig.cap='Collated index.', tidy=FALSE}
col_pal <- c("cyan4", "orange", "orangered2")

## compute the metric used for the graph of the Collated Log-Index centered around 2 (observed, bootstap sampel, credible confidence interval, linear trend)
b1 <- data.table(M_YEAR = co_index_b$M_YEAR, LCI = 2 + co_index_b$logInd - co_index_b$mean_logInd)
b2 <- data.table(M_YEAR = co_index_b[BOOTi == 0, M_YEAR], LCI= 2 + co_index_b[BOOTi == 0, logInd] - co_index_b[BOOTi == 0, mean_logInd])

lm_mod <- try(lm(LCI ~ M_YEAR, data = b2), silent=TRUE)

plot(b1, col = adjustcolor( "cyan4", alpha.f = 0.2),
      xlab = "year", ylab = expression('log '['(10)']*' Collated Index'),
      xaxt="n", type = 'n')

axis(1, at = b2$M_YEAR)

points(b2[!is.na(LCI),], type = 'l', lty=2, col="grey70")
points(b2, type = 'l', lwd=1.3, col = col_pal[1])
points(b2[!is.na(LCI),], col= col_pal[1], pch=19)
abline(h=2, lty=2)

if (class(lm_mod)[1] != "try-error"){
    points(b2$M_YEAR, as.numeric(predict(lm_mod, newdata = b2, type = "response")),
    type = 'l', col='maroon4', lwd = 1.5, lty = 1)
  }
```


##### 6 Bootstrap confidence interval

Confidence interval around the collated indices can be computed by a bootstrap method, where *n* site indices are randomly resampled *k* time to produce a distribution of annual collated indices from which confidence interval can be derived.

First we build define *k* bootstrap sample, with replacement, with the function `boot_sample()`, here we will set 200 samples, but for reliable confidence interval, this number should be set to 1000.

```{r, boot}
bootsample <- rbms::boot_sample(sindex, boot_n = 200)
```

Using the `collated_index()` function in a loop, with bootstrap sample to inform the argument `boot_ind`, we can produce *k* collated index for each year.

```{r, collated_index_boot}
co_index <- list()

## for progression bar, uncomment the following
## pb <- txtProgressBar(min = 0, max = dim(bootsample$boot_ind)[1], initial = 0, char = "*",  style = 3)

for(i in c(0,seq_len(dim(bootsample$boot_ind)[1]))){

 co_index[[i+1]] <- rbmsD::collated_index(data = sindex, s_sp = 2, sindex_value = "SINDEX", bootID=i, boot_ind= bootsample, glm_weights=TRUE, rm_zero=TRUE)

## for progression bar, uncomment the following
## setTxtProgressBar(pb, i)

}

## collate and append all the result in a data.table format
co_index <- rbindlist(lapply(co_index, FUN = "[[","col_index"))
```

Annual log indices and the average are then computed from the original sample. Similarly, annual log indices are then computed for each bootstrap sample.

```{r log_ind_boot}
co_index_b <- co_index[COL_INDEX > 0.0001 & COL_INDEX < 100000, ]
co_index_logInd <- co_index_b[BOOTi == 0, .(M_YEAR, COL_INDEX)][, log(COL_INDEX)/log(10), by = M_YEAR][, mean_logInd := mean(V1)]

## merge the mean log index with the full bootstrap dataset
data.table::setnames(co_index_logInd, "V1", "logInd"); setkey(co_index_logInd, M_YEAR); setkey(co_index_b, M_YEAR)
co_index_b <- merge(co_index_b, co_index_logInd, all.x = TRUE)

data.table::setkey(co_index_b, BOOTi, M_YEAR)
co_index_b[ , boot_logInd := log(COL_INDEX)/log(10)]
```

From the bootstrap sample, we can now derive a 95% Confidence Interval from the corresponding percentiles
```{r ind_fig_CI, fig.cap='Collated index with 95% CI.', tidy=FALSE}

b1 <- data.table(M_YEAR = co_index_b$M_YEAR, LCI = 2 + co_index_b$boot_logInd - co_index_b$mean_logInd)
b2 <- data.table(M_YEAR = co_index_b[BOOTi == 0, M_YEAR], LCI= 2 + co_index_b[BOOTi == 0, logInd] - co_index_b[BOOTi == 0, mean_logInd])
b5 <- b1[co_index_b$BOOTi != 0, quantile(LCI, 0.025, na.rm = TRUE), by = M_YEAR]
b6 <- b1[co_index_b$BOOTi != 0, quantile(LCI, 0.975, na.rm = TRUE), by = M_YEAR]
lm_mod <- try(lm(LCI ~ M_YEAR, data = b2), silent=TRUE)

## define graph axis limits and color scheme
yl <- c(floor(min(b5$V1, na.rm=TRUE)), ceiling(max(b6$V1, na.rm=TRUE)))

col_pal <- c("cyan4", "orange", "orangered2")

## draw the plot for the selected species
plot(b1, ylim = yl, col = adjustcolor( "cyan4", alpha.f = 0.2),
      xlab = "year", ylab = expression('log '['(10)']*' Collated Index'),
      xaxt="n", type = 'n')

axis(1, at = b2$M_YEAR)

segments(x0 = as.numeric(unlist(b5[,1])), y0 = as.numeric(unlist(b5[,2])),
         x1 = as.numeric(unlist(b6[,1])), y1 = as.numeric(unlist(b6[,2])),
         col = col_pal[2], lwd = 2)
points(b2[!is.na(LCI),], type = 'l', lty=2, col="grey70")
points(b2, type = 'l', lwd=1.3, col = col_pal[1])
points(b2[!is.na(LCI),], col= col_pal[1], pch=19)
abline(h=2, lty=2)

if (class(lm_mod)[1] != "try-error"){
    points(b2$M_YEAR, as.numeric(predict(lm_mod, newdata = b2, type = "response")),
    type = 'l', col='maroon4', lwd = 1.5, lty = 1)
  }
