---
title: 2. Get started with rbms - collate index
subtitle: Inputting missing counts
author: Reto Schmucki (UKCEH)
date: 28 November 2019
---
---------

From the flight curve and the observed counts, we can derive expected values for weeks or days where a site has not been monitored. Together, observed and imputed counts are used to compute abundance indices across sites. Site indices are then used to calculate annual collated indices.

> see 1. Get started with `rbms` to compute the flight curve object used below (Step 1, 2 and 3)

```{r data}
library(data.table)
library(rbms)
data(m_visit)
data(m_count)
ts_date <- rbms::ts_dwmy_table(InitYear = 2000, LastYear = 2003, WeekDay1 = 'monday')
ts_season <- rbms::ts_monit_season(ts_date, StartMonth = 4, EndMonth = 9, StartDay = 1, EndDay = NULL, CompltSeason = TRUE, Anchor = TRUE, AnchorLength = 2, AnchorLag = 2, TimeUnit = 'w')
ts_season_visit <- rbms::ts_monit_site(m_visit, ts_season)
ts_season_count <- rbms::ts_monit_count_site(ts_season_visit, m_count, sp = 2)
ts_flight_curve <- rbms::flight_curve(ts_season_count, NbrSample = 300, MinVisit = 5, MinOccur = 3, MinNbrSite = 5, MaxTrial = 4, GamFamily = 'nb', SpeedGam = FALSE, CompltSeason = TRUE, SelectYear = NULL, TimeUnit = 'w')
```

##### **4.** Impute predicted counts for missing monitoring dates

The `impute_count()` function uses the count data generated from the `ts_season_count()` function and the flight curves (ts_flight_curve$pheno) obtained from the `flight_curve()` function. The `impute_count()` function looks for the phenology available (using the nearest year) to estimate and input missing values; the extend of the search for nearest phenology can be limited by setting the YearLimit parameter. By default, this is not restricted and will look over all years available. Like other rbms functions, the imputation can be made on a weekly or daily basis ('w' or 'd').

```{r, impute}
## extract phenology data from the ts_fligh_curve list
pheno <- ts_flight_curve$pheno
## note: extraction of the pheno part is not necessary any more, as this is now done internally if not yet extract. Nevertheless the process still work and remains backward compatible.
## ts_flight_curve = ts_flight_curve would also work

impt_counts <- rbms::impute_count(ts_season_count=ts_season_count, ts_flight_curve=pheno, YearLimit= NULL, TimeUnit='w')
```

The `impute_count()` function produces a data.table that contains the original COUNT values, a series of IMPUTED_COUNT over monitoring season, TOTAL_COUNT per site and year, TOTAL_NM (e.g. the proportion of the flight curve covered by the visits) and SINDEX. The SINDEX is the site index that corresponds to the sum of both observed and imputed counts over the sampling season. If the flight curve of a specific year is missing, the `impute_count()` function uses the nearest phenology found. If none is available within the limit of years set by the YearLimit parameter, the function will return no SINDEX for that specific year.


```{r, impute_year}
impt_counts_1year <- rbms::impute_count(ts_season_count=ts_season_count, ts_flight_curve=pheno[M_YEAR != 2001, ], YearLimit= 1, TimeUnit='w')

impt_counts_0year <- rbms::impute_count(ts_season_count=ts_season_count, ts_flight_curve=pheno[M_YEAR != 2001, ], YearLimit= 0, TimeUnit='w')
```


##### **5.** Site and collated indices

From the imputed count, the site index can be calculated for each site or with a filter that will only keep sites that have been monitored at least a certain proportion of the flight curve. In this example, we set the threshold to 10%, using the MinFC parameter.

```{r, sindex, message=FALSE, warning=FALSE}
sindex <- rbms::site_index(butterfly_count = impt_counts, MinFC = 0.10)
```

With the site indices, annual collated indices can be estimated by fitting a Generalised Linear Model (GLM), where sites and years are modelled as factors. Here we also use the proportion of the flight curve sampled by the observation as a weight for the GLM. Finally, we remove all sites where the species was not observed, setting the parameter rm_zero = TRUE will facilitates model fit.

```{r, collated_index}
co_index <- collated_index(data = sindex, s_sp = 2, sindex_value = "SINDEX", glm_weights = TRUE, rm_zero = TRUE)
```

The collated index computed by the `collated_index()` function can be interpreted as the mean total butterfly count expected on a BMS transect in a given year.

```{r, collated_index_print, echo = FALSE}
co_index
```

The index can be transformed to a log(10) scale

```{r log_ind}
co_index <- co_index$col_index
co_index_b <- co_index[COL_INDEX > 0.0001 & COL_INDEX < 100000, ]
co_index_logInd <- co_index_b[BOOTi == 0, .(M_YEAR, COL_INDEX)][, log(COL_INDEX)/log(10), by = M_YEAR][, mean_logInd := mean(V1)]

## merge the mean log index with the full bootstrap dataset
data.table::setnames(co_index_logInd, "V1", "logInd"); setkey(co_index_logInd, M_YEAR); setkey(co_index_b, M_YEAR)
co_index_b <- merge(co_index_b, co_index_logInd, all.x = TRUE)
```

The log scaled indices can then be plotted with the following code; here we represent the data with time-series average being centred to two (i.e. `log10(100)`).

```{r, ind_fig, fig.cap='Collated index.', tidy=FALSE}
col_pal <- c("cyan4", "orange", "orangered2")

## compute the metric used for the graph of the Collated Log-Index centred around 2 (observed, bootstrap sample, credible confidence interval, linear trend)
b1 <- data.table(M_YEAR = co_index_b$M_YEAR, LCI = 2 + co_index_b$logInd - co_index_b$mean_logInd)
b2 <- data.table(M_YEAR = co_index_b[BOOTi == 0, M_YEAR], LCI= 2 + co_index_b[BOOTi == 0, logInd] - co_index_b[BOOTi == 0, mean_logInd])

lm_mod <- try(lm(LCI ~ M_YEAR, data = b2), silent=TRUE)

plot(b1, col = adjustcolor( "cyan4", alpha.f = 0.2),
      xlab = "year", ylab = expression('log '['(10)']*' Collated Index'),
      xaxt="n", type = 'n')

axis(1, at = b2$M_YEAR)

points(b2[!is.na(LCI),], type = 'l', lty=2, col="grey70")
points(b2, type = 'l', lwd=1.3, col = col_pal[1])
points(b2[!is.na(LCI),], col= col_pal[1], pch=19)
abline(h=2, lty=2)

if (class(lm_mod)[1] != "try-error"){
    points(b2$M_YEAR, as.numeric(predict(lm_mod, newdata = b2, type = "response")),
    type = 'l', col='maroon4', lwd = 1.5, lty = 1)
  }
```


##### **6.** Bootstrap confidence interval

The confidence interval around the collated indices can be computed from a bootstrap sample, where *n* site indices are randomly resampled (with replacement) *k* time to produce a distribution of the annual collated indices. From this distribution, we can derive the confidence intervals around the collated indices.
We first define *k* bootstrap sample using the function boot_sample(). Here we set *k* to 200 samples, but for a reliable confidence interval, *k* should be at least 1000. For reproducibility, we use set.seed() to generate a repeatable random sample


```{r, boot}
set.seed(218795)
bootsample <- rbms::boot_sample(sindex, boot_n = 200)
```

Using the `collated_index()` function in a loop, with bootstrap samples informing the argument `boot_ind`, we can now compute *k* collated indies over the entire time-series.

```{r, collated_index_boot, message=FALSE, warning=FALSE}
co_index <- list()

## for progression bar, uncomment the following
## pb <- txtProgressBar(min = 0, max = dim(bootsample$boot_ind)[1], initial = 0, char = "*",  style = 3)

for(i in c(0,seq_len(dim(bootsample$boot_ind)[1]))){

 co_index[[i+1]] <- rbms::collated_index(data = sindex, s_sp = 2, sindex_value = "SINDEX", bootID=i, boot_ind= bootsample, glm_weights=TRUE, rm_zero=TRUE)

## for progression bar, uncomment the following
## setTxtProgressBar(pb, i)

}

## collate and append all the result in a data.table object
co_index <- rbindlist(lapply(co_index, FUN = "[[","col_index"))
```

Annual log indices, as well as their average, are then computed from the original sample. Similarly, we compute annual log indices for each bootstrap sample.

```{r log_ind_boot}
co_index_b <- co_index[COL_INDEX > 0.0001 & COL_INDEX < 100000, ]
co_index_logInd <- co_index_b[BOOTi == 0, .(M_YEAR, COL_INDEX)][, log(COL_INDEX)/log(10), by = M_YEAR][, mean_logInd := mean(V1)]

## merge the mean log index with the full bootstrap dataset
data.table::setnames(co_index_logInd, "V1", "logInd"); setkey(co_index_logInd, M_YEAR); setkey(co_index_b, M_YEAR)
co_index_b <- merge(co_index_b, co_index_logInd, all.x = TRUE)

data.table::setkey(co_index_b, BOOTi, M_YEAR)
co_index_b[ , boot_logInd := log(COL_INDEX)/log(10)]
```

From the bootstrap samples, we can derive a 95% Confidence Interval, using the corresponding percentiles (i.e., 0.025 and 0.975).

```{r ind_fig_CI, fig.cap='Collated index with 95% CI.', tidy=FALSE}

b1 <- data.table(M_YEAR = co_index_b$M_YEAR, LCI = 2 + co_index_b$boot_logInd - co_index_b$mean_logInd)
b2 <- data.table(M_YEAR = co_index_b[BOOTi == 0, M_YEAR], LCI= 2 + co_index_b[BOOTi == 0, logInd] - co_index_b[BOOTi == 0, mean_logInd])
b5 <- b1[co_index_b$BOOTi != 0, quantile(LCI, 0.025, na.rm = TRUE), by = M_YEAR]
b6 <- b1[co_index_b$BOOTi != 0, quantile(LCI, 0.975, na.rm = TRUE), by = M_YEAR]
lm_mod <- try(lm(LCI ~ M_YEAR, data = b2), silent=TRUE)

## define graph axis limits and color scheme
yl <- c(floor(min(b5$V1, na.rm=TRUE)), ceiling(max(b6$V1, na.rm=TRUE)))

col_pal <- c("cyan4", "orange", "orangered2")

## draw the plot for the selected species
plot(b1, ylim = yl, col = adjustcolor( "cyan4", alpha.f = 0.2),
      xlab = "year", ylab = expression('log '['(10)']*' Collated Index'),
      xaxt="n", type = 'n')

axis(1, at = b2$M_YEAR)

segments(x0 = as.numeric(unlist(b5[,1])), y0 = as.numeric(unlist(b5[,2])),
         x1 = as.numeric(unlist(b6[,1])), y1 = as.numeric(unlist(b6[,2])),
         col = col_pal[2], lwd = 2)
points(b2[!is.na(LCI),], type = 'l', lty=2, col="grey70")
points(b2, type = 'l', lwd=1.3, col = col_pal[1])
points(b2[!is.na(LCI),], col= col_pal[1], pch=19)
abline(h=2, lty=2)

if (class(lm_mod)[1] != "try-error"){
    points(b2$M_YEAR, as.numeric(predict(lm_mod, newdata = b2, type = "response")),
    type = 'l', col='maroon4', lwd = 1.5, lty = 1)
  }
